Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2020-03-11 paper/method.tex
minor edits to middle sections

For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
Curtis \etal\ (2020, in prep) combined effective temperature measurements for
nearby, unreddened field stars in benchmark samples, including FGK stars
characterized with high-resolution optical spectroscopy \citep{brewer2016}, M
dwarfs characterized with low-resolution optical and near-infrared
spectroscopy \citep{mann2015}, and K and M dwarfs characterized with
interferometry and bolometric flux analyses \citep{boyajian2012}.
This empirical color--temperature relation is valid over the color range $0.55
< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
3070$~K.
The dispersion about the relation implies a high precision of 50~K.
These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
at all temperatures\footnote{The color--temperature relation is described in
detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
\etal\ (2020, in prep).}.
The equation we used to calculate photometric temperatures from Gaia \gcolor\
color is a seventh-order polynomial with coefficients given in table
\ref{tab:coeffs}.
% \ref{tab:coeffs}.
% \begin{equation}
%     \mathrm{T_{eff}} = a + bC cC^2 + dC^3
%     eC^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
% % Jason's updated parameters:
% % −416.585
% % 39780.0
% % −84190.5
% % 85203.9
% % −48225.9
% % 15598.5
% % −2694.76
% % 192.865
%     % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
%     % 301C^6 - 26C^7,
% % 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
% %           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
% %           301.13205924630165, -25.923997443169355]
% \label{eq:curtis}
% \end{equation}
% where C is \gaia\ \gcolor and the coefficients are provided in table
% \ref{coeffs}.
\begin{table}[h!]
  \begin{center}
      \caption{
          Coefficient values for a 7th-order polynomial used to estimate
      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
      prep).}
    \label{tab:coeffs}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
      Exponent & Coefficient  \\
      \hline
      $0$ & -416.585 \\
      $1$ & 39780.0  \\
      $2$ & -84190.5 \\
      $3$ & 85203.9  \\
      $4$ & -48225.9 \\
      $5$ & 15598.5  \\
      $6$ & -2694.76 \\
      $7$ & 192.865  \\
    \end{tabular}
  \end{center}
\end{table}

Photometric binaries and subgiants were removed from the sample by applying
cuts to the color-magnitude diagram (CMD), shown in figure
\ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and photometric binaries (shown
as the curved dashed line in figure \ref{fig:age_gradient}).
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 4th magnitude
in \gaia\ G-band.

The dwarf stars in the \mct\ sample are shown on a \gaia\ color-magnitude
diagram (CMD) in figure \ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2011}, and are therefore more likely to have measurable
rotation periods at old ages, the latter scenario seems more likely.
An alternative explanation could be that the rotation periods of the oldest
early M dwarfs are so long that they are not measurable with Kepler data.
Ground-based rotation period measurements of mid and late M dwarfs indicate
that there is an upper limit to the rotation periods of M dwarfs of around 140
days \citep{newton2016, newton2017, newton2018}, which is much longer than the
longest rotation periods measured in the \mct\ sample (around 70 days).
For this reason, it is likely that slowly rotating M dwarfs are missing from
the \mct\ sample and this could be another reason for the lack of old M
dwarfs.

The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
Since this relation predicts that the oldest stars in the \mct\ sample are
late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
and early-M dwarfs.
This is a typical feature of empirically calibrated gyrochronology models:
since there are no (or at least very few) old M dwarfs with rotation periods,
the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    % \includegraphics[width=1\textwidth]{age_gradient_combo}
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate stellar velocities.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
ovariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided below) were removed from the sample.
% Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
% and because latitude is correlated with stellar mass (lower mass stars are
% older and tend to be preferentially located at high $b$, although this trends
% is greatly reduced for stars cooler than 5000 K), we found that excluding
% stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
% stellar mass.
% 15\degrees\ was found to be an adequate compromise between maintaining a close
% relationship between \vb\ and \vz\ and retaining a sample large enough to
% provide meaningful results.
% Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
% trends in the data, and intermediate cuts did not significantly change the
% results.
% We also found that the anti-correlation between stellar mass and $b$ is
% negligable for stars cooler than 5000 K.
% For this reason, and because the rotational evolution of old, G stars does not
% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
% dwarfs.

>>>>

For this work, we used the precise \textit{Gaia} DR2 photometric color,
$G_{\rm BP} - G_{\rm RP}$, to estimate \teff\ for the Kepler rotators.
Curtis \etal\ (2020, in prep) combined effective temperature measurements for
nearby, unreddened field stars in benchmark samples, including FGK stars
characterized with high-resolution optical spectroscopy \citep{brewer2016}, M
dwarfs characterized with low-resolution optical and near-infrared
spectroscopy \citep{mann2015}, and K and M dwarfs characterized with
interferometry and bolometric flux analyses \citep{boyajian2012}.
This empirical color--temperature relation is valid over the color range $0.55
< (G_{\rm BP} - G_{\rm RP})_0 < 3.20$, corresponding to $6470 < T_{\rm eff} <
3070$~K.
The dispersion about the relation implies a high precision of 50~K.
These benchmark data enable us to accurately estimate \teff\ for cool dwarfs
\citep[\eg][]{rabus2019}, and allows us to correct for interstellar reddening
at all temperatures\footnote{The color--temperature relation is described in
detail in the Appendix of, and the formula is provided in Table 4 of, Curtis
\etal\ (2020, in prep).}.
The equation we used to calculate photometric temperatures from Gaia \gcolor\
color is a seventh-order polynomial with coefficients given in table
\ref{tab:coeffs}.
% \ref{tab:coeffs}.
% \begin{equation}
%     \mathrm{T_{eff}} = a + bC cC^2 + dC^3
%     eC^4 + 15598.5C^5 -2694.76C^6 + 192.865C^7,
% % Jason's updated parameters:
% % −416.585
% % 39780.0
% % −84190.5
% % 85203.9
% % −48225.9
% % 15598.5
% % −2694.76
% % 192.865
%     % \mathrm{T_{eff}} = 8960 -4802C + 1931C^2 -2446C^3 + 2669C^4 - 1324C^5 +
%     % 301C^6 - 26C^7,
% % 8959.8112335205078, -4801.5566310882568, 1931.4756631851196,
% %           -2445.9980716705322, 2669.0248055458069, -1324.0671020746231,
% %           301.13205924630165, -25.923997443169355]
% \label{eq:curtis}
% \end{equation}
% where C is \gaia\ \gcolor and the coefficients are provided in table
% \ref{coeffs}.
\begin{table}[h!]
  \begin{center}
      \caption{
          Coefficient values for the 7th-order polynomial used to estimate
      \teff\ from \Gaia\ \gcolor\ color, calibrated in Curtis \etal\ (2020, in
      prep).}
    \label{tab:coeffs}
    \begin{tabular}{l|c} % <-- Alignments: 1st column left and 2nd middle, with vertical lines in between
        (\gcolor\ ) exponent & Coefficient  \\
      \hline
      $0$ & -416.585 \\
      $1$ & 39780.0  \\
      $2$ & -84190.5 \\
      $3$ & 85203.9  \\
      $4$ & -48225.9 \\
      $5$ & 15598.5  \\
      $6$ & -2694.76 \\
      $7$ & 192.865  \\
    \end{tabular}
  \end{center}
\end{table}

Photometric binaries and subgiants were removed from the \mct\ sample by
applying cuts to the color-magnitude diagram (CMD), shown in figure
\ref{fig:age_gradient}.
A 6th-order polynomial was fit to the main sequence and raised by 0.27 dex to
approximate the division between single stars and photometric binaries (shown
as the curved dashed line in figure \ref{fig:age_gradient}).
All stars above this line were removed from the sample.
Subgiants were also removed by eliminating stars brighter than 4th magnitude
in \gaia\ G-band.

The rotation periods of the dwarf stars in the \mct\ sample are shown on a
\gaia\ color-magnitude diagram (CMD) in the top panel of figure
\ref{fig:age_gradient}.
In the bottom panel, the stars are colored by their gyrochronal age,
calculated using the \citet{angus2019} gyrochronology relation.
The stars with old gyrochronal ages, plotted in purple hues, predominantly lie
along the upper edge of the MS, where stellar evolution models predict old
stars to be, however the majority of these `old' stars are bluer than \gcolor\
$\sim$ 1.5 dex.
The lack of gyrochronologically old M dwarfs suggests that either old M dwarfs
are missing from the \mct\ catalog, or the \citet{angus2019} gyrochronology
relation under-predicts the ages of low-mass stars.
Given that lower-mass stars stay active for longer than higher-mass stars
\citep[\eg][]{west2011, kiman2019}, and are therefore more likely to have
measurable rotation periods at old ages, the latter scenario seems likely.
However, it is also possible that the rotation periods of the oldest early M
dwarfs are so long that they are not measurable with Kepler data.
Ground-based rotation period measurements of mid and late M dwarfs indicate
that there is an upper limit to the rotation periods of {\it late} M dwarfs of
around 140 days \citep{newton2016, newton2017, newton2018}, which is much
longer than the longest rotation periods measured in the \mct\ sample (around
70 days).
The apparent lack of old ages for M dwarfs in figure \ref{fig:age_gradient}
may be caused by a combination of ages being underestimated by a poorly
calibrated model, and rotation period detection bias.
% For this reason, it is likely that slowly rotating M dwarfs are missing from
% the \mct\ sample and this could be another reason for the lack of old M
% dwarfs.
The \citet{angus2019} gyrochronology relation is a simple polynomial model,
fit to the period-color relation of Praesepe.
% Since this relation predicts that the oldest stars in the \mct\ sample are
% late-G and early-K dwarfs, it is probably under-predicting the ages of late-K
% and early-M dwarfs.
Inaccuracies are a typical feature of empirically calibrated gyrochronology
models: since there are no (or at least very few) old M dwarfs with rotation
periods, the models are poorly calibrated for these stars.
\begin{figure}
  \caption{
      Top: de-reddened MS \kepler\ stars with \mct\ rotation periods, plotted
    on a \gaia\ CMD.
    We removed photometric binaries and subgiants from the sample by excluding
    stars above the dashed lines.
    Bottom: a zoom-in of the top panel, with stars colored by their
    gyrochronal age \citep{angus2019}, instead of their rotation period.
    A general age gradient is visible across the main sequence.
    Since the \citet{angus2019} relation predicts that the oldest stars in
    the \mct\ sample are late-G and early-K dwarfs, it is probably
    under-predicting the ages of late-K and early-M dwarfs.
}
  \centering
    % \includegraphics[width=1\textwidth]{age_gradient_combo}
    \includegraphics[width=1\textwidth]{CMD_cuts_double}
\label{fig:age_gradient}
\end{figure}

The {\tt Pyia} \citep{price-whelan_2018} and {\tt astropy} \citep{astropy2013,
astropy2018} {\it Python} packages were used to calculate velocities for the
\mct\ sample.
{\tt Pyia} calculates velocity samples from the full \gaia\ uncertainty
covariance matrix via Monte Carlo sampling, thereby accounting for the
covariances between \gaia\ positions, parallaxes and proper motions.
Stars with negative parallaxes, parallax signal-to-noise ratios less than 10,
stars fainter than 16th magnitude, stars with absolute \vb\ uncertainties
greater than 1 \kms\, and stars with galactic latitudes greater than
15\degrees\ (justification provided below) were removed from the sample.
% Because \vb\ is only a close approximation to \vz\ at low galactic latitudes,
% and because latitude is correlated with stellar mass (lower mass stars are
% older and tend to be preferentially located at high $b$, although this trends
% is greatly reduced for stars cooler than 5000 K), we found that excluding
% stars with $b > 15$\degrees\ reduced the number of \vb\ outliers at low
% stellar mass.
% 15\degrees\ was found to be an adequate compromise between maintaining a close
% relationship between \vb\ and \vz\ and retaining a sample large enough to
% provide meaningful results.
% Reducing this cut to 10\degrees\ resulted in a sample size too small to reveal
% trends in the data, and intermediate cuts did not significantly change the
% results.
% We also found that the anti-correlation between stellar mass and $b$ is
% negligable for stars cooler than 5000 K.
% For this reason, and because the rotational evolution of old, G stars does not
% follow a simple Skumanich-like spin-down law \citep{vansaders2016}, we
% restricted our sample to temperatures between \tmin\ and 5000 K, \ie\ K
% dwarfs.

### 2020-03-11 paper/results.tex
minor edits to middle sections

To explore the relationship between rotation period, \teff\ and velocity
dispersion, we first removed high and low velocity outliers from the \mct\
sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remain the
same without sigma-clipping.
% We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
% 3500 K to avoid biases caused by the selection function at the faint, cool
% end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with rotation periods shorter than the bulk of
periods, since this area of the period-\teff\ diagram is sparsely populated.
We removed rapid rotators by cutting out stars with gyrochronal ages less than
0.5 Gyr, because a 0.5 Gyr gyrochrone\footnote{A gyrochrone is a
gyrochronological isochrone, or a line of constant age in period-\teff, or
period-color space.} traces the bottom edge of the main population of rotation
periods.
After these cuts, 6879 stars were included in the sample.

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age, as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones they would follow lines of constant velocity
dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a more
flattened, or even inverted relationship between rotation period and effective
temperature than these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citep{spada2019} model shown in white.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduces the apparent
flattening and inversion in the rotation period-\teff/mass relations.

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period decreases with increasing
mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for K-dwarfs.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.

>>>>

To explore the relationship between rotation period, effective temperature
(\teff ) and velocity dispersion, we first removed high and low velocity
outliers from the \mct\ sample by performing 3$\sigma$ sigma-clipping on the
\vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, although this did not change the overall
trends in the data.
% We also limited the sample to temperatures in the range 6000 K $<$ \teff\ $<$
% 3500 K to avoid biases caused by the selection function at the faint, cool
% end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with rotation periods shorter than the main
population of periods, since this area of the period-\teff\ diagram is
sparsely populated.
We removed these rapid rotators by cutting out stars with gyrochronal ages
less than 0.5 Gyr \citep[based on the][gyro-model]{angus2019}, because a 0.5
Gyr gyrochrone\footnote{A gyrochrone is a gyrochronological isochrone, or a
line of constant age in period-\teff, or period-color space.} traces the
bottom edge of the main population of rotation periods.
After these cuts, around 13,000 stars out of the original $\sim$34,000 were
included in the sample.

The top panel of figure \ref{fig:vplot} shows rotation period versus effective
temperature for the \mct\ sample, coloured by the standard deviation of their
(\vb) velocities, where \sigmavb\ was calculated for groups of stars over a
grid in $\log_{10}$(period) and temperature.
If we assume that mass dependent heating does not strongly affect this sample
and \vb\ at low galactic latitudes is an unbiased tracer of \vz, then \vb\
velocity dispersion can be interpreted as an age proxy, and stars plotted in a
similar color in figure \ref{fig:vplot} are similar ages.
We discuss this assumption further in section
\ref{sec:mass-dependent-heating}.
\begin{figure}
  \caption{
    Top: Rotation period vs effective temperature for stars in the \mct\
    sample, colored by the velocity dispersions of stars calculated over a
    grid in $\log_{10}$(period) and \teff.
    Black lines show gyrochrones from a gyrochronology model that projects the
    rotation-color relation of
    Praesepe to longer rotation periods over time \citep{angus2019}.
    These gyrochrones do not appear to reflect the evolution of field stars at
    long rotation periods/old ages because they do not trace lines of constant
    velocity dispersion.
    Gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr in both top
    and bottom panels.
    Bottom: Same as top panel with rotation period vs {\it mass}
    \citep[from][]{berger2020}.
    White lines show gyrochrones from a model that includes mass and
    age-dependent angular momentum transport between the core and envelope
    \citep{spada2019}.
    Qualitatively, these gyrochrones reflect the evolution of field
    stars at long rotation periods/old ages: they trace lines of constant
    velocity dispersion by reproducing periods of `stalled' rotational
    evolution for K-dwarfs.
}
  \centering
    \includegraphics[width=1\textwidth]{main_figure}
\label{fig:vplot}
\end{figure}
Overall, figure \ref{fig:vplot} shows that velocity dispersion increases with
rotation period across all temperatures, implying that rotation period
increases with age, as expected.
This result is insensitive to the choice of bin position and size.
Black lines show gyrochrones from the \citet{angus2019} gyrochronology model,
which projects the rotation-color relation of Praesepe to longer rotation
periods over time.
These gyrochrones are plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
At the youngest ages, these gyrochrones describe the data well: the palest
yellow (youngest) stars with the lowest velocity dispersions all fall close to
the 0.5 Gyr gyrochrone.
% Lines of constant age (isochrones) appear to follow the shape of the
% Praesepe-based gyrochronology model (black dotted line) at young ages.
% However, at older ages it appears that the relation between rotation period
% and \teff\ flattens out, until eventually rotation period {\it decreases} with
% decreasing effective temperature at a given age.
However, although the 0.5 Gyr and 1 Gyr gyrochrones also trace constant
velocity dispersion/age among the field stars, by 1.5 Gyr the gyrochrones
start to {\it cross} different velocity dispersion regimes.
For example, the 1.5 Gyr gyrochrone lies on top of stars with velocity
dispersions of around 10-11 kms$^{-1}$ at 5000-5500K and stars with $\sim$15
\kms\ velocity dispersions at 4000-4500K.
The gyrochrones older than 1.5 Gyr also cross a range of velocity dispersions.
If these were true isochrones they would follow lines of constant velocity
dispersion.
At ages older than around 1 Gyr, it appears that gyrochrones should have a
more flattened, or even inverted, shape in rotation period-\teff\ space than
these Praesepe-based models.
% For late K-dwarfs older than around 1 Gyr, it appears that lower-mass stars
% rotate more rapidly than higher-mass stars.
% This is the opposite of the trend that is observed in young open clusters such
% as Praesepe and the Hyades.

The bottom panel of figure \ref{fig:vplot} shows velocity dispersion as a
function of rotation period and {\it mass}, \citep[from][]{berger2020}, with
gyrochrones from the \citep{spada2019} model shown in white.
These gyrochrones are also plotted at 0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr.
Each point plotted in the top panel also appears in the bottom panel with the
same color.
Because velocity dispersion was calculated in bins of \teff, not mass, bin
outlines are clearly visible in the top panel but appear smeared-out in the
bottom panel.
In the bottom panel of figure \ref{fig:vplot}, the \citet{spada2019} models
{\it do} trace lines of constant velocity dispersion, and reproduce the trends
in the data at all ages.
These models qualitatively agree with the data and reproduce the apparent
flattening and inversion in the rotation period-\teff/mass relations.

The results shown in figure \ref{fig:vplot} indicate that stars of spectral
type ranging from late G to late K ($\sim$5500-3500 K) follow a braking law
that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.
The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages \citep[younger than around 1 Gyr but still old enough to be on
the main sequence and have transitioned from the `I' sequence to the `C'
sequence ][]{barnes2003}, stellar rotation period {\it decreases} with {\it
increasing} mass.
This is likely because lower-mass stars with deeper convection zones have
stronger magnetic fields, larger Alfv\'en radii and therefore experience
greater angular momentum loss rate \citep[\eg][]{schatzman1962, parker1970,
kawaler1988, charbonneau2010}.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
This aligns with the current assumptions about stellar spin-down, dynamo
theory, and the gyrochronology paradigm that has been in place for decades
\citep[\eg][]{skumanich1972, noyes1984, kawaler1988, barnes2003, angus2019}.
According to the data presented in figure \ref{fig:vplot} however, at
intermediate ages, the rotation periods of K dwarfs appear {\it constant} with
mass, and at late ages rotation period {\it increases} with {\it increaasing}
mass.
The interpretation of this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly, and this effect is strongest for late K-dwarfs with
effective temperatures of $\sim$4000-4500K and masses $\sim$0.5-0.7 M$_\odot$.
